---
title: "Ireland Housing - Data Exploration"
author: Marcos Cavalcante
output: github_document
---
First step is to install and load the necessary libraries.

```{r loadlib, echo=T, results='hide', message=F, warning=F}
packages <- c("tidyverse","haven", "devtools", "dplyr")

if(sum(as.numeric(!packages %in% installed.packages())) != 0){
  installer <- packages[!packages %in% installed.packages()]
  for(i in 1:length(installer)) {
    install.packages(installer, dependencies = T)
    break()
  }
  sapply(packages, require, character = T) 
} else {
  sapply(packages, require, character = T) 
}

devtools::install_github("ropensci/skimr")
library(skimr)

```


## Data Exploration

At this next stage, the dataset will be explored further with regards to the data values it holds.
Exploration around the lower and upper boundaries of each numeric attribute as well as its histogram distribution will be done.


Let's start by using the skim library to have a glimpse of the dataset.

```{r}

dataset_directory <- "../../datasets/"
dataset_filename <- paste(dataset_directory, "ireland_houses_filtered.csv", sep="")

ireland_houses <- read.csv(file = dataset_filename ) # Load the dataset

skim(ireland_houses)
```


Glimpse of the dataset without the latitude and longitude attributes:

```{r}
skim(ireland_houses) %>%
  filter(!skim_variable %in% c("latitude", "longitude"))
```

## Understanding the data

The data already provides really good insights. For example: 

* The mean _price_ of houses is €601,026 and the most expensive house in the dataset is worth €15 million euro.

* The mean _size_ of a house in $sq^2$ is 130 $sq^2$, and the biggest house has 6300 $sq^2$.

* There appears to be some outliers as well, for instance: house with 30 bedrooms and 28 bathrooms
