---
title: "Ireland Housing - Data Exploration"
author: Marcos Cavalcante
output: github_document
---
First step is to install and load the necessary libraries.

```{r Libraries Installation, loadlib, echo=T, results='hide', message=F, warning=F}
packages <- c("tidyverse", "haven", "devtools", "dplyr", 
              "ggplot2", "gapminder", "patchwork", "ggridges", "PerformanceAnalytics")

if(sum(as.numeric(!packages %in% installed.packages())) != 0){
  installer <- packages[!packages %in% installed.packages()]
  for(i in 1:length(installer)) {
    install.packages(installer, dependencies = T)
    break()
  }
  sapply(packages, require, character = T) 
} else {
  sapply(packages, require, character = T) 
}

devtools::install_github("ropensci/skimr")
library(skimr)

```


## Data Exploration

At this next stage, the output from the cleaning step will be used in to further explore the variables, to make it explicitly, the dataset __ireland_houses_cleaned.csv__ is going to be used.


For continuous variables, exploration around the lower and upper boundaries of each variable, their histogram distribution and box-plots will be created.

With regards to the categorical variables, visualizing their distribution will be quite important so that the frequency of the different categories can be seen in the dataset.

Let's start by using the skim library to have a glimpse of the dataset.

```{r Load Dataset}

dataset_directory <- "../../datasets/"
dataset_filename <- paste(dataset_directory, "ireland_houses_cleaned.csv", sep="")

ireland_houses <- read.csv(file = dataset_filename ) # Load the dataset

options(scipen = 999) # turn off scientific notation
skim(ireland_houses)
```


## Understanding the data

The summary output above already provides great insights into the dataset. For example:

### Categorical Variables

#### Address

* There seems to be about 8,881 different addresses, and the total number of observations is 78,034.

#### Property Type

* **Property Type** holds only 9 different values, they are: *Apartment*, *Bungalow*, *Detached*, *Duplex*, *End of Terrace*, *Semi Detached*, *Terrace*, *Town House* and **NA**.

#### Category

* It is present in all observations and holds 2 possible values: *Buy* and *New Houses*.

#### Town

* This variable holds the neighborhoods in Ireland, it has 68 different values. This shows that not all the cities / neighborhoods from Ireland are in this dataset - this is expected as there may be no houses on sale in those places.

### Continuous Variables

#### Price
* The mean _price_ of houses is around €600,000 and the *most expensive* house in the dataset is worth €15 million euro.

* **Percentiles**: It can be seen on the table that very few, nearly 0% of the houses have a price of up to €40,000, 25% prices of up to €300,000, 50% prices of up to €415,000, 75% prices of up to €650,000 and no houses are over €15 million euro. Understanding the percentiles is quite helpful as it shows how the house prices are distributes over the observations.

* **Standard Deviation**: The standard deviation in the price variable is quite high, €656.000, which indicates that there is a great price range.

#### Property Size

* The mean _size_ of a house in $sq^2$ is 130 $sq^2$, and the biggest house has 6300 $sq^2$.

* **Percentiles**: In the first percentile it shows that there are some observations with under 1$sq^2$; 25% of the houses have up to 75$sq^2$; 50% of the observations are under 102$sq^2$, 75% are under 144$sq^2$ and there are no houses bigger than 6109$sq^2$.

* **Standard Deviation**: The standard deviation in the *property size* variable is quite high, 160$sq^2$. What this tells us is that there is a large range of house sizes, as a comparison, the standard deviation is greater than 75% of the observations.

#### Bathrooms

* The **bathrooms** variable shows some interesting behaviour, where the mean is about 2 bathrooms per house and the maximum number of bathrooms in a house is 28.

#### Bedrooms

* Similar to the *bathrooms* variable, **bedrooms** also show an intersting behaviour, the mean is 3 bedrooms per house and the maximum number of bedrooms is 30.


## Frequency graphs

Let's start the data exploration by plotting frequency graphs in order to understand how the observations are distributed over different variables, for example: *property type*, *property size* and *house price*.


### Property Type Bar Chart

The bar chart below shows the observations are distributed over the property types. It is clear that the majority of the houses are terrace and are followed by semi-detached, detached and apartments respectively.

The graph also shows that some houses have no value for property type and therefore have missing data.

```{r Property Type Bar Chart }
ggplot(data = ireland_houses) +
  geom_bar(
    width = 0.4,
    mapping = aes(fill = as.factor(propertyType), 
                  x = factor(propertyType,
                      labels = c("NA", 
                                 "Apartment", 
                                 "Bungalow", 
                                 "Detached", 
                                 "Duplex", 
                                 "End of Terrace", 
                                 "Semi-Detached", 
                                 "Terrace", 
                                 "Town House"
  )))) +
  scale_y_continuous(breaks = seq(0, 27000, 2000)) +
  labs(title = "Property Types Count", 
       x = "Property Type",
       y = "Count") +
  coord_flip() +
  theme(legend.position="none")
```


### Property Type by Proper



```{r}
chart.Correlation((ireland_houses[2:4]), histogram = TRUE)
```

### Property Type by Price Bivariate graph

After understanding how the observations are distributed across the 16 valid property types. It can clearly be seen that *Terrace*, *Semi-Detached* and *Detached* houses roughly represent over 75% of the houses.

The next graph displays how prices and property prices are related.

```{r Property Type by Price Bivariate graph}

# propertyTypeFactor = factor(propertyType,labels = c(
#                                  "NA", 
#                                  "Apartment", 
#                                  "Bungalow", 
#                                  "Detached", 
#                                  "Duplex", 
#                                  "End of Terrace", 
#                                  "Semi-Detached", 
#                                  "Terrace", 
#                                  "Town House"))


ggplot(data = ireland_houses) + 
  geom_density_ridges(aes(x = price, 
                          y = propertyType, 
                          fill = propertyType)) +
  scale_x_continuous(name = "price", limits = c(0, 1000000)) +
  theme_ridges() + 
  labs(title = "Property Type by Price", x = "Price", y = "Property Type") +
  theme(legend.position = "none")

```

### House Price

In order to better visualize how the observations distribute across the price range, let's choose the range from €0 to €2 million euro as it will allow us to well visualise where most of the observations fall.

```{r House Prices Histogram, warning=FALSE}
ggplot(ireland_houses, aes(x = price, colour = propertyType, fill = propertyType)) +
  geom_freqpoly(
    bins=30,
    lwd = 1,
    linetype = 2
  ) +
  labs(
    title = "Frequency of House Prices in Ireland",
    x = "Price",
    y = "Observations",
    color = "Property Type"
  ) +
  xlim(0, 2000000) +
  scale_fill_brewer() 
```


### Property Type
For this graph, for visualization sake, let's pick the house which have property size smaller than 750 $sq^2$

```{r Property Size Histogram, warning=FALSE}

ggplot(ireland_houses, aes(x = propertySize, colour = propertyType, fill = propertyType)) +
  geom_freqpoly(
    bins=30,
    lwd = 1,
    linetype = 2
  ) +
  labs(
    title = "Frequency of Property Size in Ireland",
    x = "Property size in square meters",
    y = "Observations",
    color = "Property Type"
  ) +
  xlim(0, 500) +
  scale_fill_brewer() 
```
